---
title: "Regression Models - Course Project"
author: "Giovanni Picardi"
date: "30 gennaio 2015"
output: pdf_document
---

# Executive summary

Dataset mtcars has been explored trying to investigate how the variable `am`, an indicator variable of manual transmission, is related to `mpg` (miles per US gallon). Considering `am` alone shows a higher mean `mpg` for manual than for automatic transmission, but other variables have a stronger correlation with the outcome: weight `wt`, displacement `disp` and gross horsepower `hp`. Fitting and analyzing linear models with `am` and one of these variables (centered) as predictors, possibly considering interaction, leads to the conlcusion that no absolute difference of `mpg` can be attributed to transmission type for the average values of the other predictor (intercept), but only a different slope favouring automatic transmission, although in both cases transmission type has a too strong correlation with the other predictor, making the conclusion too tied to the linear model hypothesis. On the other hand a simple model considering `am` and `hp`, with transmission type equally distributed in the range of the `hp` values in the dataset, reveals a higher expected `mpg` for manual transmission for a given `hp`.

# Exploratory data analysis

The datasets has no missing values, all of the `r dim(mtcars)[2]` variables are numeric and they are related to a range of different motorcars (`r dim(mtcars)[1]` models).
A comparison of boxplots of the variable `mpg` for automatic and manual transmission (**Figure 1**) shows that the mean mileage per gallon is higher for manual transmission (hypothesis tested in the following section).
The outcome `mpg` shows the following correlation coefficients with the other variables:
```{r, echo=FALSE}
data(mtcars)
cmpg <- round(cor(mtcars$mpg, mtcars),3)
cmpg <- cmpg[,order(abs(cmpg), decreasing=T)]
cmpg[2:11]
```
The relationship between pairs of variables, outcome inlcuded, can be visually examined with `pairs(mtcars)` (**Figure 2**): the more "linearly" related to `mpg` are the weight `wt`, the displacement `disp` and the horse power `hp`.
The relationship between these four variables with the outcome `mpg` is detailed in **Figure 3**, highlighting the different transmission types. 

# Modelling
Fitting a linear model for `mpg` with the single predictor `am` (i.e.: `mpg ~ am`) allows to compare the mean `mpg` for automatic and manual transmission:
```{r, echo=FALSE}
mc <- transform(mtcars, am=factor(am, labels=c("Automatic","Manual")))
fitam <- lm(mpg ~ am, mc)
summary(fitam)$coefficients[2,]
```
and to say with appropriate confidence that manual transmission cars are expected to show a mean mileage `r round(coef(fitam)[2],2)` miles per US gallon higher than automatic transmission cars.

Linear models where `am` appears toghether with one of `wt`, `disp` and `hp`, **centered on their respective means**, were analyzed to verify the effects of other variables and the explicative power of `am` on `mpg`.
  
The ANOVA for the nested linear models `mpg ~ am` and `mpg ~ am + wt` would suggest that the inclusion of `wt` is highly desirable, but the `am` coefficient shows a too high p-value, so no effect on expected `mpg` for mean `wt` can be attributed to `am` with appropriate confidence; moreover the ANOVA for the nested models `mpg ~ wt` and `mpg ~ wt + am` would on the contrary suggest that `wt` alone is sufficient: this is in agreement with the distribution of the different type of transmission with respect to weight (see **Figure 3**). The inclusion of the interaction term between `wt` and `am` whould leads to a more explicative model `mpg ~ wt*am` whose coefficients are
```{r, echo=FALSE}
mc <- transform(mc, wt=wt-mean(wt), disp=disp-mean(disp), hp = hp-mean(hp))
fitamwt <- lm(mpg ~ wt*am, mc)
summary(fitamwt)$coefficients
```
that explains a lot of variance (adjusted $R^2=$ `r round(summary(fitamwt)$adj.r.squared,4)`), and states, again, that no absolute difference is to be expected in `mpg` between automatic and manual transmission at the mean weight (intercept), but that the expected decrease in `mpg` for a 1000 lb increment in weight is `r -round(summary(fitamwt)$coefficients[4,1],2)` miles/gallon larger in magnitude for manual than for automatic transmission. Diagnostics for the model are in **Figure 4**.

Repeating the same analysis with `disp` instead of `wt` leads to very similar conclusions: no effect on expected `mpg` for mean `disp` can be attributed to `am` with appropriate confidence and `disp` alone seems sufficient in explaining `mpg` (see again **Figure 3**). The inclusion of the interaction term leads to a more explicative model `mpg ~ disp*am` whose coefficients are
```{r, echo=FALSE}
fitamdisp <- lm(mpg ~ disp*am, mc)
summary(fitamdisp)$coefficients
```
with adjusted $R^2=$ `r round(summary(fitamdisp)$adj.r.squared,4)`, that states, again, that no absolute difference is to be expected in `mpg` between automatic and manual transmission at the mean displacement (intercept), but that the expected decrease in `mpg` for a 1 cu. in. increment in displacement is `r -round(summary(fitamdisp)$coefficients[4,1],2)` miles/gallon larger in magnitude for manual than for automatic transmission. Diagnostics for the model are in **Figure 5**.

The ANOVA for the models `mpg ~ am` and `mpg ~ am+hp`, compared to the ANOVA for the models `mpg ~ hp` and `mpg ~ hp+am` shows that the model with two predictor is always preferable to the ones with a single predictor. The fitting of the model `mpg ~ am+hp` leads to the coefficients
```{r, echo=FALSE}
fitamhp <- lm(mpg ~ am+hp, mc)
summary(fitamhp)$coefficients
```
with adjusted $R^2=$ `r round(summary(fitamhp)$adj.r.squared,4)`, that allows to draw the conclusion that manual transmission gives rise, for fixed `hp`, to an expected `mpg` `r round(summary(fitamhp)$coefficients[2,1],2)` miles/gallon larger than automatic transmission. Diagnostics for the model are in **Figure 6**, and show no pattern in residuals and the best Q-Q plot: this is the soundest model.

# Results

Manual transmission is better on average for MPG: the expected MPG increment is `r round(coef(fitam)[2],2)` miles/gallon, with 95% confidence interval (`r round(confint(fitam)["amManual",],2)`).
Manual transmission is better on average for fixed gross horse power: the expected MPG increment is `r round(summary(fitamhp)$coefficients[2,1],2)` miles/gallon, with 95% confidence interval (`r round(confint(fitamhp)[2,],2)`).


# Appendix

```{r, echo=FALSE, fig.height=4}
par(cex=0.8)
boxplot(mpg ~ as.factor(c("Automatic","Manual")[mtcars$am+1]), data=mtcars,
        ylab="Miles (US) per gallon", xlab="Transmission",
        main="Figure 1 - Miles per gallon and Transmission")
```
    

```{r, echo=FALSE}
pairs(mtcars, cex=0.6, main="Figure 2 - Relationship between couples of variables",
      upper.panel=panel.smooth)
```

```{r, echo=FALSE}
par(mfrow=c(2,2), mar=c(5,4,1,1), oma=c(0,0,1,0))
with(mtcars, {
  plot(wt, mpg, col=am+1)
  legend("topright", legend=c("Automatic", "Manual"), pch=1, col=c(1,2))
  plot(disp, mpg, col=am+1)
  plot(hp, mpg, col=am+1)
  plot(cyl, mpg, col=am+1)
  mtext(outer=TRUE, text="Figure 3 - Main variables, type of transmission and outcome", font=2)
})
```

```{r, echo=FALSE}
par(mfrow=c(2,2), mar=c(5,4,2,1), oma=c(0,0,2,0))
plot(fitamwt, sub.caption="Figure 4 - Diagnostics for model mpg ~ wt*am")
```

```{r, echo=FALSE}
par(mfrow=c(2,2), mar=c(5,4,2,1), oma=c(0,0,2,0))
plot(fitamdisp, sub.caption="Figure 5 - Diagnostics for model mpg ~ disp*am")
```

```{r, echo=FALSE}
par(mfrow=c(2,2), mar=c(5,4,2,1), oma=c(0,0,2,0))
plot(fitamhp, sub.caption="Figure 6 - Diagnostics for model mpg ~ hp+am")
```